---
title: "R Summary Sheet"
subtitle: BIO 46, Winter 2018
output: 
  html_document:
    highlight: pygments
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: true
---

```{r, include = FALSE}

# Remove XX from code chunk output display
knitr::opts_chunk$set(comment = NA)

# Cache chunk output for faster rendering
knitr::opts_chunk$set(cache=TRUE)

# Do not display warning messages
knitr::opts_chunk$set(warning=FALSE)

# Increase the width of the console for printing output
options(width = 200)

```

I usually include this code at the top of scripts so that R doesn't automatically read text as factors.

```{r, warning = FALSE, message = FALSE, include = FALSE}
# Read in text as text and not factors
options(stringsAsFactors = FALSE)

```


# Data manipulation

## Loading Data

The following data tables are available for analysis:

**Table Descriptions**

+ **sites**:  Location and canopy cover of Jasper Ridge survey sites.
+ **lichens**: Info about lichens collected at Jasper Ridge, including outcomes of growth and infection experiments. 
+ **fungal_isolates**: Growth rates and identities of fungi isolated from lichens.
+ **infection_prevalence**: Abundance of *E. prunastri* and infection prevalence at survey sites. 
+ **interaction_experiment_byplate**: Outcome of fungal isolate interaction experiment for each plate.
+ **interaction_experiment_bycompetitor**: Outcome of fungal isolate interaction experiment for each competitor on a plate.
+ **env_loggers_summary**: Temperature and humidity summary data for each site.
+ **infection_experiment**: Results of infection experiment.


**Table locations**

| Table name | Google sheet key | GitHub filename |
|------------|------------------|----------------|
| sites | 1LJj8BAngXFjBTvpOJv_0WXV_OwTXtEHk00p_WKsMuJ8 | BIO46_2018_sample_sites.csv |
| lichens | 11d3-k-3OaIOfm8EfneUiNGiumfPBdMBHMViaTbVtzwY | lichens.csv |
| fungal_isolates | 16ku-VRUHMZHz5XgHtD2ESDhdem401sBnbbPEEVeYtrg | |
| infection_prevalence | 1iJ1bM4qGkL8M068wJAr52BU801hHo6HldQClU75FCMc | infection_prevalence.csv |
| interaction_experiment_byplate| 1ty7Wz0oiEIFJcfSCy8FFIVd3eyiwI0iMc0AyGVdFrpY | interaction_experiment_byplate.csv |
| interaction_experiment_bycompetitor | 16YxUnW2o1WfYYab5IUNTzxFVcrKVvwerbR7pGLJzHZQ | interaction_experiment_bycompetitor.csv |
| env_loggers_summary | 1ktSM1PYOnazM663Fg_bdTYRPA8yUlHH2VYFriHFleEo | env_loggers_summary.csv |
| infection_experiment | 1RwIhDQNeAWcLI9ZlrlpekDWQ0cjPM_UkBgb5FFBDiM8 | infection_experiment.csv |

Three ways to load these data:

**1) From the Google Drive sheets**

```{r message = FALSE}
# Load package
library(googlesheets)

# Authorize googlesheets to access you Google Drive
gs_auth()

# Register the lichens data table
lichens_gs <- gs_key("11d3-k-3OaIOfm8EfneUiNGiumfPBdMBHMViaTbVtzwY")

# Read the lichens_gs googlesheet into R
lichens <- gs_read(lichens_gs)
```

**2) From the GitHub repository**

```{r}
# Define the URL of the data repository
github_url <- "https://raw.githubusercontent.com/jescoyle/BIO46/master/2018/Data/"

# Read in the csv file using the url and the specific file name
# The following reads in three different tables
interactions <- read.csv(file.path(github_url, "interaction_experiment_bycompetitor.csv"))
sites <- read.csv(file.path(github_url, "BIO46_2018_sample_sites.csv"))
infection <- read.csv(file.path(github_url, "infection_prevalence.csv"))

```

**3) From a file on your computer**

First download the files to your BIO46 folder.

```{r eval = FALSE}
# Set the working directory
setwd("C:/Users/your_username/Documents/BIO46")

# Read in a csv file that you downloaded to that directory
read.csv("lichens.csv")

```

## Manipulating Data {.tabset}

Functions for manipulating dataframes are in the `dplyr` and `tidyr` packages. Be sure to load them once before using these functions.

```{r message = FALSE, warning = FALSE}
# Load the packages
library(dplyr)
library(tidyr)
```


### Subsetting

Use `select` to choose specific columns and `filter` to choose specific rows based on the values in the columns. In the code below, the resulting dataframe is saved to an object named `shrub_inf` that can be used later.

```{r}
# Show the site and lichen number of lichens on shrubs that have more than 0 ascomata (i.e. infected)
shrub_inf <- lichens %>%
  filter(Ascomata > 0 & Substrate_type == "shrub") %>%
  select(LichenID, Site, Ascomata)


shrub_inf
```

### Summarizing

Summarize values in columnes with the `summarise` function. Use `group_by` to specify the columns used to group data before calculating summary statistics. In the code examples below, the resulting dataframes are saved to an objects named `substrate_summary` and `mean_growth` that can be used later.

```{r message = FALSE, warning = FALSE}
# For lichens on different substrates,
# Count the number of lichens found 
#  and calculate mean and maximum number of ascomata
substrate_summary <- lichens %>%
  group_by(Substrate_type) %>%
  summarise(Num_lichens  = n(),
            Ascomata = mean(Ascomata),
            Ascomata_max = max(Ascomata)
  )

substrate_summary
```

```{r}
# Calculate the mean radial growth of each isolate growing solo or in co-culture
mean_growth <- interactions %>%
  group_by(IsolateID, Culture_type) %>%
  summarise(Radius_mm_day7 = mean(Radius_mm_day7),
            Radius_mm_day14 = mean(Radius_mm_day14)
  )

mean_growth
```

### Stacking columns

To combine multiple columns into the same column, use the `gather` function. `key =` is the name of a new column that will identify which column the data originally came from. `value = ` is the name of a new column that will contain the values that were originally in separate columns. 

```{r}
# Put the radius measurements from day 7 and day 14 into a single column (named Radius_mm)
mean_growth_long <- mean_growth %>%
  gather(key = Measurement_day, value = Radius_mm, Radius_mm_day7, Radius_mm_day14)

mean_growth_long
```


### Calculating new columns

Add new columns based on calculations with other columns using `mutate`.

```{r}
# Add a new column to lichens that indicates whether it is infected (ascomata > 0)
lichens <- lichens %>%
  mutate(Infected = Ascomata > 0) 

# Display the LichenID, Ascomata and Infected columns
lichens %>%
    select(LichenID, Ascomata, Infected)
```

```{r}
# Add columns to mean_growth_long that indicate the number of days
# and then calculate growth rate (mm per day)
mean_growth_long <- mean_growth_long %>%
  mutate(Days = ifelse(Measurement_day == "Radius_mm_day7", 7, 14), 
         Growth_mm_day = Radius_mm / Days
  )
         
mean_growth_long
```

### Combining tables

Merge the columns of two dataframes using a `join` function. `left_join` keeps all of the rows of the left dataframe. `right_join` keeps all of the rows of the right dataframe. `full_join` keeps all rows in both dataframes. The function will use the column names that are the same in both dataframes to match up the rows.

```{r message = FALSE}
# Select some columns from the site data
sites_light <- sites %>%
  select(Site, Canopy_openness_pct)

# Add infection prevalence data to the abbreviated site data
inf_sites <- left_join(sites_light, infection)

inf_sites
```

# Data visualization {.tabset}

We will use functions from the `ggplot2` package for data visualization.

```{r}
# Load the package
library(ggplot2)
```

## Scatterplot

A scatterplot shows how two variable relate to one another. Each point corresponds to a row in a dataframe. You can include a linear regression line using `geom_smooth(method = "lm")` or a logistic regression line using `geom_smooth(method = "glm")`.

```{r warning = FALSE, message = FALSE}
# Plot the number of ascomata versus a lichen's initial mass
lichens %>%
  ggplot(aes(x = Mass_mg_init, y = Ascomata)) +
    geom_point() +
    labs(x = "Initial mass (mg)",
         y = "Number of ascomata"
    )
```

```{r}
# Create the same plot with a linear regression line.
lichens %>%
  ggplot(aes(x = Mass_mg_init, y = Ascomata)) +
    geom_point() +
    geom_smooth(method = "lm",
                se = TRUE,
                level = 0.95) +
    labs(x = "Initial mass (mg)",
         y = "Number of ascomata"
    )
```

```{r warning = FALSE}
# Create a plot of infectedion status versus initial mass
# including a logistic regression line
lichens %>%
  mutate(Infected = as.numeric(Infected)) %>% # convert to 0 or 1
  ggplot(aes(x = Mass_mg_init, y = Infected)) +
    geom_point() +
    geom_smooth(method = "glm",
                method.args = list(family = "binomial"),
                se = TRUE,
                level = 0.95) +
    labs(x = "Initial mass (mg)",
         y = "Infected"
    )

```

## Histogram

A histogram shows the frequency of observed X values.

```{r}
# Create a histogram that counts how many lichens have different numbers of ascomata
lichens %>%
  ggplot(aes(x = Ascomata)) +
    geom_histogram(binwidth = 5, boundary = 5) +
    labs(x = "Num. ascomata",
         y = "Number of lichens"
    )
```

## Boxplot

A boxplot shows the distribution (median and quartiles) of Y values in different X categories.

The following boxplot uses data from the `mean_growth_long` dataframe created above in the Manipulating data -> Stacking columns section. Here are the first 10 rows...

```{r echo = FALSE}
knitr::kable(mean_growth_long[1:10,], digits = 2)
```

```{r warning = FALSE, message = FALSE}
# Boxplot of the growth rates of fungal isolates growing solo and in co-culture
mean_growth_long %>%
  ggplot(aes(x = Culture_type, y = Growth_mm_day)) +
    geom_boxplot() +
    labs(x = "Culture type",
         y = "Growth rate (mm / day)"
    )
```


## Barchart

Barcharts can show the number of observations in different categories.

```{r}
# Plot the number of infected vs. uninfected lichens on trees vs. shrubs
lichens %>%
  ggplot(aes(x = Substrate_type)) +
    geom_bar(aes(fill = Infected),
             position = "dodge" # remove this if you want the bars stacked
    ) +
    labs(x = "Host plant",
         y = "Num. lichens"
    )

```

Barcharts can also show the mean and standard devation of observations in different categories.

```{r warning = FALSE, message = FALSE}
# Calculate mean and sd of growth on day 7 across taxa 
rad7_summary <- interactions %>%
  group_by(TaxonID, Culture_type) %>%
  summarise(Radius_mm = mean(Radius_mm_day7),
            Radius_sd = sd(Radius_mm_day7)
  )


# Plot the mean and sd of different taxa 
# growing solo versus in co-culture
rad7_summary %>%
  ggplot(aes(x = TaxonID, y = Radius_mm, fill = Culture_type)) +
    geom_col(position = "dodge") +
    geom_errorbar(aes(ymin = Radius_mm - Radius_sd,
                      ymax = Radius_mm + Radius_sd),
                  position = position_dodge(width = 0.9),
                  width = 0.5
    ) +
    labs(x = "",
         y = "Colony radius day 7 (mm)",
         fill = "Culture type"
    ) +
    theme(axis.text.x = element_text(angle=90, 
                                     vjust = 0.5, 
                                     hjust = 1
                        )
    )
```

## Saving Figures

To save a figure, click the "Export" button in the Plots window of RStudio. 

Alternatively, save the figure directly from your code by running the following immediately after generating a figure. Change the extension to .pdf or .svg if you prefer a vector format image.

```{r eval = FALSE}
# Save the last plot as my_file_name.png in the current working directory.
ggsave("my_file_name.png")

```


# Statistics 

## Y categorical {.tabset}

We'll use the `lichens` dataframe downloaded above to demonstrate analyses for when Y is categorical. Here are the first 10 rows...

```{r echo = FALSE}
knitr::kable(lichens[1:10,])
```

### X categorical

To apply the following tests, first create a contingeny table of the data that tabulates how many observations are from each combination of X and Y categories. These tests evaluate whether the counts in the rows of the contingency table are different from one another. 

```{r contingency-table}
# Count the number of infected vs. uninfected lichens on trees vs shrubs
lichen_count <- table(lichens$Substrate_type, lichens$Infected)
  
# Display table 
lichen_count

```

#### Chi-square test

Use this test when you have many observations and the expected number of observations in each cell of the table is greater than five. This test is not valid when you have few observations.

```{r}
# Test whether infection frequency is different between trees and shrubs
chisq.test(lichen_count)

```

#### Fisher's exact test

Use this test when you have few observations and when you know ahead of time how many observations are in each category of X and each category of Y (e.g. the marginal sums are fixed). Statisticians recommend not using this test and instead using unconditional tests (see [Lydersen et al. 2009](http://dx.doi.org/10.1002/sim.3531)).

```{r fisher-test}
# Fisher's exact test with a two-sided hypothesis
fisher.test(lichen_count)

```

#### Unconditional exact tests

Use this test when you have few observations and when you do not necessarily know ahead of time how many observations are in each category of X and each category of Y (e.g. the marginal sums are random). 

```{r}
# Load the package containing the function
library(Exact)
```

```{r barnard-test, message = FALSE}
# Barnard's test used when the number of observations in each category of X is known
# X categories should be in the rows of the table, otherwise, specify, cond.row = FALSE
# method = "CSM" specifies Barnard's method, but you can use other options
barnard_test <- exact.test(lichen_count, 
                           model = "binomial", 
                           method = "CSM", 
                           cond.row = TRUE, 
                           to.plot = FALSE)

# Print test results
barnard_test
```

```{r}
# Boschloo's test used when the number of observations in each category is unknown for X or Y
boschloo_test <- exact.test(lichen_count, 
                            model = "multinomial", 
                            method = "boschloo")

# Print test results
boschloo_test
```

### X continuous

#### Logistic regression

Use this form of logistic regression when the dependent response variable has two outcomes (e.g. False / True, Absent / Present).

```{r}
# Fit a logistic regression model with:
# Y = whether a lichen is infected
# X = the initial mass of the lichen
mod_infVmass <- glm(Infected ~ Mass_mg_init, 
                   data = lichens, 
                   family = binomial(link = "logit")
)

# View the results
summary(mod_infVmass)

# Exponentiate the mode coeffifients (intercept and slope)
# to see the effect of mass on the odds of being infected
exp(coef(mod_infVmass))

```


## Y continuous {.tabset}

### X categorical 

We'll use the `interactions` dataframe downloaded above to demonstrate ANOVA and t-tests. Here are the first 10 rows... 

```{r echo = FALSE}
knitr::kable(interactions[1:10,])
```

#### ANOVA

Use ANOVA when you want to test whether a normally-distributed dependent variable differs among categories. ANOVA assumes that the variance within groups is the same across groups.

```{r}
# Replace unknown (NA) taxon names with their isolate ID (e.g. S1-A)
int_noNAtaxa <- interactions %>%
  mutate(TaxonID = ifelse(is.na(TaxonID), IsolateID, TaxonID))

# ANOVA testing whether day 7 radius differs among taxa
# Fit model
mod_rad7Vtaxa <- lm(Radius_mm_day7 ~ factor(TaxonID), data = int_noNAtaxa)

# Perform ANOVA
anova(mod_rad7Vtaxa)
```

**Post-hoc t-test**

ANOVA only tells you whether the categories are different from one another (variance between categories >  variance within categories). If you want to know which categories differ from one another, use Tukey's Honest Significant Difference test after performing an ANOVA to determine whether the pairs of categories have different means. 

```{r}
# First fit an AOV (analysis of variance) object
aov_rad7Vtaxa <- aov(mod_rad7Vtaxa)

# Then computed Tukey's HSD
TukeyHSD(aov_rad7Vtaxa)
```

#### t-test

Use a t-test whether you want to test whether the mean of a normally-distributed dependent  variable differs between two categories.

The arguments in the `t.test` function can be used to specify whether the variances should be assumed equal among groups (Student's t-test, `var.equal = TRUE`) or unequal (Welch's t-test, `var.equal = FALSE`). The `paired` arguments determines whether you want an independent sample or dependent sample test. Use a dependent sample test (`paired = TRUE`) when you have measured  the same individuals in both categories. 

We'll use the `interactions` data to demonstrate and test whether the radius measured on day 7 differed from the radius measured on day 14.

```{r}
# Student's t-test, two-sided hypothesis
t.test(interactions$Radius_mm_day14, 
       interactions$Radius_mm_day7,
       var.equal = TRUE)

```

```{r}
# Student's t-test, one-sided hypothesis
t.test(interactions$Radius_mm_day14, 
       interactions$Radius_mm_day7,
       var.equal = TRUE,
       alternative = "greater")

```
```{r}
# Welch's t-test (unequal variance), one-sided hypothesis
t.test(interactions$Radius_mm_day14, 
       interactions$Radius_mm_day7,
       var.equal = FALSE,
       alternative = "greater")
```

Since measurements are not independent- we measured the same plates on day 7 and day 14, a paired test is more appropriate. Note that the two columns must have "individuals" in the same order.

```{r}
# Paired Welch's t-test treating plates as individuals
t.test(interactions$Radius_mm_day14, 
       interactions$Radius_mm_day7,
       var.equal = FALSE,
       alternative = "greater",
       paired = TRUE)
```

### X continuous

We'll use the `inf_sites` dataframe created above in the Manipulating Data section to demonstrate fitting linear regression models. Here are the first 20 rows...

```{r echo = FALSE}
knitr::kable(inf_sites[1:20,])
```


#### Linear regression

Use normal linear regression when the dependent variable is continuous and (mostly) normally distributed.

```{r}
# Linear regression of number of E prunastri vs. canopy openness
mod_eprunVcanopy <- lm(Num_Eprunastri ~ Canopy_openness_pct, data = inf_sites)

# View the results
summary(mod_eprunVcanopy)

```

#### Count (Poisson) regression

Poisson regression is useful when the dependent variable is a discrete count. The number of *E. prunastri* is a count, not actually continuous and since there are many zeros and low counts (see histogram below) we may want to use Poisson regression instead. 

```{r}
# Histogram to view the distribution of Num_Eprunastri
inf_sites %>%
  ggplot(aes(x = Num_Eprunastri)) +
    geom_histogram(binwidth = 5, boundary = 5)

```

```{r}
# Fit a Poisson regression
mod_eprunVcanopy_pois <- glm(Num_Eprunastri ~ Canopy_openness_pct, 
                             data = inf_sites,
                             family = "poisson"
)

# View the results
summary(mod_eprunVcanopy_pois)

```

```{r}
# Create a scatterplot
# fitted Poisson regression line in blue
# fitted linear regression line in red
inf_sites %>%
  ggplot(aes(x = Canopy_openness_pct, y = Num_Eprunastri)) +
    geom_point() +
    geom_smooth(method = "glm",
                method.args = list(family = "poisson"),
                color = "blue") +
    geom_smooth(method = "lm",
                color = "red")
```





