---
title: "Linear Models and Community Similarity"
author: "BIO 46, Winter 2017, Stanford University"
date: "Week 7: February 21, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = '', eval=F)
```

### Lesson Goals

The goal of this lesson is to introduce how to fit various linear models in R, including normal linear regression, logistic regression, and other GLMs. We will also be learning how to calculate community similarity metrics and statistically analyze variation in these metrics. We will be working with a data table that has been created for you by merging three data tables from our class database: 

1. *carbon_utilization*: a table derived from measurment of the ecoplates, giving the amount of each carbon substrate utilized at each time step by each microbial community sample
2. *bacteria_plates*: a table indicated where each microbial community was sampled, including the tree, lichen species (*Ramalina menziesii* and *Evernia prunastri*), thallus location (internal vs. external), and infection status (healthy vs infected)
3. *ecoplates*: a table linking the carbon\_utilization and bacteria\_plates tables

The data table you will be working with is called *merged_final_carbon_utilization_data.csv* and it contains carbon substrate utilization from the last date that each plate was measured, together with identifying information about the where the microbial community in each sample came from (e.g. which lichen, internal vs. external, healthy vs. infected). The code that was used to derived this data table can be found in the R Lesson 5 supplement.

You should download the merged data file from Canvas and save it to the Data folder for our class. 

As you work through this lesson, be sure to run each line of code in the script (which corresponds to the code in the lesson). Remember- lines that begin with `#` are comments that tell you what the next line of code is supposed to do.
If you can't tell how a whole line of code worked, try highlighting small pieces of the line of code and running them separately to see what they do. You only need to run each line of code once- any time that you assign an object a name, R will remember that object until you re-use that name for a different object. If you forget what objects you have created and named you can look in the Environment section of RStudio, which lists a description of each object you have created and assigned a name.

In the **YOUR TURN** sections you should type code that accomplishes the required task. When you complete the lesson, save your R script and upload it to the R Lesson 4 assignment on Canvas. 


**YOUR TURN**: Modify the code below so that you can read the two data tables into your R work session.

```{r}
# Change the working directory
setwd(  )

# Read the data file merged_final_carbon_utilization_data.csv into a dataframe named analysis_data
analysis_data = read.csv(  )

```

Click on the analysis_data object in the Environment window in R Studio to view the data that is included in the table.


### Linear Regression

Linear regression is used to evaluate the relationship between a continuous predictor variable and continuous response variable whose residuals after regression follow a normal distribution. For example, we could use linear regression to examine whether there is a relationship between the amount of a substrate used and the mass of lichen from which the microbial community was extracted.

The first step in any data analysis is to visualize the relationship between the two variables. For two continuous variables we will use an x-y plot. Let's examine mannitol since this is one of the primary carbohydrates available in a lichen thallus (green algal photobionts produce ribitol which the mycobionts then covert into mannitol).

```{r}
# Make an x-y plot of the amount of mannitol used vs. mass of lichen
plot(D.Mannitol ~ Lichen_mass_mg, data = analysis_data)
```

We might have expected a positive relationship, but instead it doesn't look like the ammount of mannitol utilized is related to the mass of lichen from which the microbial commuinty was extracted. Let's fit a regression line to verify this. Linear regression in R in conducted using the same linear model function we used for ANOVA: `lm()`. The syntax is: `lm(y_var ~ x_var, data = dataframe_with variables)`.

```{r}
# Fit a linear model to mannitol with a single linear predictor, lichen_mass_mg
man_mass_mod = lm(D.Mannitol ~ Lichen_mass_mg, data=analysis_data)

# View the fitted model
summary(man_mass_mod)
```

In the summary table, the coefficient estimates are giving the y-intercept (Intercept) and slope (Lichen\_mass\_mg) of the line of best fit. The 'slope' or estimated effect of Lichen\_mass\_mg is -0.001 meaning that if 1 mg more lichen was used there wuld be 0.001 units less mannitol. The t statistic and its associated P-value are testing whether this coefficient is different from 0. A P-value of 0. shows that it is not. Therefore there is not effect of lichen mass on mannitol utilization.

we can plot the line of best fit on the existing figure using the function `abline()`. When used with a linear model, `abline` extracts the model coeffients and draws a line with the correct y-intercept and slope.

```{r}
# Add the estimated line of best fit to the figure
abline(man_mass_mod, col='red')
```


**YOUR TURN**: Evaluate whether the amount of mannitol utilized is related to bacterial abundance (measured as CFUs per mg thallus tissue).

1. Use your notes from class to calculate CFUs per mg of lichen thallus based on the following columns in `analysis_data`: `Lichen_mass_mg`, `Frac_extract`, `Init_vol_ml `, `Dilution_factor`, `Plate_vol_ml`. Add this variable as a new column in `analysis_data` nameed `CFU_mg`.
2. Make an x-y plot showing the relationship between CFUs and mannitol utilization.
3. Fit a linear regression testing whether this relationship is statistically significant.
4. Add the estimated line of best fit to the x-y plot.



### Generalized Linear Models

#### Logistic Regression

What if we are not interested in the amount of mannitol utilized, but whether manitol was utilized or not? In this case, the response variable is binary, so we need to use logistic regression to determine whether the probability that mannitol was used is related to the mass of lichen from which the microbial community was extracted.

Currently each column named after a carbon source is showing a number proportional to the amount of that carbon source that was used. We need to convert these measurements to a binary indicated whether the substrate was used (TRUE) or not (FALSE). Usually this is done by selecting only those measurements that occur in the top X % of the data. We will use a threshold of 50%. That is, we will say that a substrate was used if the amount used is in the top 50% of measurements recorded for that substrate. 

```{r}
# Make a new dataframe that is the same as analysis_data, but will have binary data on substrate utilization
binary_data = analysis_data

# Select only columns corresponding to carbon substrates
substrate_cols = 9:39

# Go through each column of the substrates
for(i in substrate_cols){
  
  # Select the data in this column
  this_col = binary_data[,i]
  
  # Find the value below which 50% of the data occur
  # na.rm = T tells the function to ignore NA values
  cutoff = quantile(this_col, 0.5, na.rm = T)
  
  # Assign TRUE to all values above the cutoff and FALSE to values below it
  new_vals = this_col > cutoff
  
  # Assign FALSE to all values that are NA because these also did not have any substrate utilized
  new_vals[is.na(new_vals)] = FALSE
  
  # Put these new values back into the binary_data
  binary_data[,i] = new_vals
}

# Examine the binary_data column for mannitol
binary_data$D.Mannitol

```

**YOUR TURN**: Fill in the code below to plot the relationship between lichen mass and whether mannitol was utilized.

```{r}
# Plot mannitol usage vs. lichen mass
plot( ~  , data = )

```

Logistic regression models the probability that an event is observed as a function of a continuous variable. In our case we are modeling the probability that mannitol was utilized as a function of the amount of lichen used to extract microbial communities. Logitistic regression is a type of generalized linear model (GLM)- a linear model where the response variable is not normally distributed. In R, GLMs are fit using the function `glm()`. The syntax is similar, but includes an arguement `family = ` that gives the distribution of the response variable. In addition, the way the response variable is given on the right side of the formula changes a little depending on the type of response variable. For logistic regression with a binary response variable, we can just supply the response variable as it is (`TRUE`/`FALSE`) and specify binomial distribution with the logit link function (specified as `family = binomial(link = logit)`).


```{r}
# Fit a logistic regression model
man_glm = glm(D.Mannitol ~ Lichen_mass_mg, data = binary_data, family = binomial(link=logit))

# View the estimated model
summary(man_glm)

```

The coefficient estimate for `Lichen_mass_mg` is -0.0084, which is its effect on the log odds of mannitol being utilized. Log odds are difficult to think about, so we can convert this coeficient estimate to its effect on the odd by exponentiating.

```{r}
# Exponentiate the coefficents to see effect on odds
exp(coef(man_glm))

```

Now we see that the odds of mannitol being utilized decrease by ~ 1% (1 - 0.99) for every mg increase in lichen mass. Wald's z statistic and its associated P-value are also given in the coefficients table output. This statistic tests whether the estimated effect differs frmo zero, and here we see that it doesn't (P = 0.326). Finally, we can visulaize the effect of lichen mass on the probability that mannitol is utilized by plotting the predicted probabilities from the fitted model.

```{r}
# Make a vector of x values (lichen masses) at which you want to estimate the probability
mass_range = range(analysis_data$Lichen_mass_mg, na.rm = T) # Find min and max masses
x_vals = seq(mass_range[1], mass_range[2], length.out = 100) # Make sequence of 100 values between min and max masses

# Predict probability using fitted model at each x value
new_data = data.frame(Lichen_mass_mg = x_vals) # Make a dataframe with the same predictors that are in the model
y_vals = predict(man_glm, new_data, type='response') # type = 'response' tells the function to predict probabilities rather than log odds

# Add predicted points to graph as a red line
lines(x_vals, y_vals, col='red')

```

#### Poisson model for count data

Another common application of GLMs is when we are interested in modeling a response variable that measures a count. For example, suppose we are interested in whether the number of carbon substrates used is related to the lichen's mass. The Poisson distribution might be an appropriate error distribution to model this response variable. We can fit a GLM with Poisson distributed errors by changing the `family` argument to `family = poisson(link = log)`.

Before fitting the model we first need to count the number of substrates used in each sample.

```{r}
# Count the number of TRUES in each row for columns corresponding to carbon substrates
rowSums(binary_data[ , substrate_cols])

# Assign this count to a new column names in binary_data
binary_data$Num_substrates = rowSums(binary_data[ , substrate_cols])

# How many samples used 0, 1, 2, ... etc substrates?
substrate_counts = table(binary_data$Num_substrates)
plot(substrate_counts, xlab='Num. substrates', ylab='Num. samples')
```

Now let's visualize the data and fit the model to see whether the number of substrates utilized is affected by the mass of lichen from which the microbial community was extracted.

```{r}
# Plot number of substrates utilized vs. lichen mass
plot(Num_substrates ~ Lichen_mass_mg, data = binary_data)

# Fit a GLM with Poisson errors
subs_glm = glm(Num_substrates ~ Lichen_mass_mg, data = binary_data, family = poisson(link = 'log'))

# View the model results
summary(subs_glm)

# Predict the number of substrates utilized using fitted model at each x value
y_vals = predict(subs_glm, new_data, type='response')

# Add a red line with the predicted values
lines(x_vals, y_vals, col='red')

```

The model output is the same as in the logistic regression model, except this time the coefficient estimates are showing the effect of the predictor variable on the log of the number of substrates used (because we used a log link). To get the effect of lichen mass on the actual number of substrates we would need to exponentiate the coefficients as we did before. Regardless, the P-value on the z-statistic is 0.128, showing that there is no effect of lichen mass on the number of substrates utilized. One thing to also note is that the residual deviance is 1023 which is nearly 10 times greated than the 127 degrees of freedom in the model. This indicates that there is overdispersion and that the Poisson distribution is not a good fit for these data. This is probably because we have so many samples with 0 substrates utilized. If we were going to use this analysis in a project we should probably choose a different distribution that can account for this structure of so many zeros in the data.

In the last three models we have been using a continuous response variable (lichen mass), but GLMs can also be fit with categorical predictors. Let's fit one final model evaluating whether the probability that mannitol is utilized differs between internal and external microbiral communities. 

```{r}
# Make table of how many internal vs. external samples mannitol was utilized in
table(binary_data$Thallus_location, binary_data$D.Mannitol)

# Fit a GLM testing whether location affects the odds that mannitol will be utilized
man_loc_mod = glm(D.Mannitol ~ Thallus_location, data = binary_data, family = binomial(link = 'logit'))

# View the model results
summary(man_loc_mod)

# View the effect on the odds (rather than log odds)
exp(coef(man_loc_mod))

```

The model results are showing that the odds of an internal microbial community utilizing mannitol are 9.7% of those of an external microbial community utilizing mannitol and that this effect is significant with a P-value of $1.2\times10^{-6}$ (based on Wald's z-statistic). Note that we could have calculated this odds ratio ourselves using the values in the table:

```{r}
# Make table of how many internal vs. external samples mannitol was utilized in
table(binary_data$Thallus_location, binary_data$D.Mannitol)

odds_external = 20/22
odds_internal = 8/91
odds_internal / odds_external
```

**YOUR TURN**: Evaluate whether the probability of *any* substrate being utilized is affected by the mass of lichen from which the microbial community was extracted.

1. Make a variable that indicates whether any substrate was utilized. Add this variable as a new column in `binary_data` called `C_used`.
2. Make an x-y plot visualizing the relationship between lichen mass and whether any substrate was used.
3. Fit a GLM modeling this relationship.
4. Add a red line showing the model prediction.



### Measuring Community Similarity

In the previous sections of this lesson we examined whether measures of microbial community function (e.g. mannitol utilization) and diversity (e.g. number of substrates utilized) were related to the amount of lichen from which communities were extracted (lichen mass) or the location from which communities were extracted (internal vs. external). In this section we will examine how the composition of microbial communities differs between samples by calculating dissimilarity metrics based on carbon substrate utilization.

We will use functions from the `vegan` package to calculate dissimilarity metrics. You will need to install the package `vegan`. To install a package in R Studio you can click the 'Packages' tab in the upper right window, then click 'Install' and type 'vegan' into the line marked 'Packages'. 

The function `vegdist()` in `vegan` allows us to calculate many different measures of dissimilarity. 

```{r}
# Load the functions in the vegan package
library(vegan)

# Display the help file on vegdist to veiw the different dissimilarity measures that are available
?vegdist

```

There is a large body of literature on the properties of different similarity metrics and chosing the appropriate metric for a given applications. For demonstration purposes we will only go through one example: bray-curtis dissimilarity calculated on the amounts of each carbon substrate utilized. The `vegdist` function will compare each pair of samples (rows) in `analysis_data` and calculate the bray-curtis dissimilarity between them. 

Let's compare the microbial community composition among internal and external communities extracted from the three *Evernia prunastri* collected on Tree 1. We can use the `grep()` function to search for 'W17-T1-' in the Plate IDs and then only use data from these rows.


```{r}
# Make a subset of the data corresponding to only data from Tree 1
tree1_rows = grep('W17-T1-', analysis_data$PlateID)
tree1_data = analysis_data[tree1_rows, ]

# NOtice that the last two rows correspond to L4, which was Ramalina menziesii, so let's remove those
tree1_data = tree1_data[1:6,]

# Make a new dataframe that only includes the columns used for comparing community composition
comm_data = tree1_data[,substrate_cols]

# Add row names that show which sample each row is from
rownames(comm_data) = tree1_data$PlateID

# Replace NA values with 0
comm_data[is.na(comm_data)] = 0

# Calculate bray-curtis dissimilarity between all pairs of samples
bc_dist = vegdist(comm_data, method = 'bray')

# The warning said that some samples don't have any substrates utilized
# Let's see which sample this was: it was L3-P2
rowSums(comm_data)

# Examine the dissimilarities: notice that all samples are maximally dissimilar to L3-P2 (value = 1)
bc_dist
```

Are samples from the same location (internal or external) more similar to one another than samples from different locations? We can visualize this questios by sorting the dissimilarities into those calculated between samples from the same location, versus samples from different locations. We will do this using the `dist()` function on the vector of Thallus locations. It will return a distance martix that is the same as `bc_dist` except that it will be 0 when comparing two samples from the same location (e.g. internal-internal) and 1 when comparing an internal sample to an external sample.

```{r}
# Make a distance matrix comparing whether samples come from the same (0) or different (1) thallus locations
loc_dist = dist(as.numeric(tree1_data$Thallus_location))
loc_dist

# Sort the dissimilarities into categories based on loc_dist
plot(loc_dist, bc_dist)
```

From the plot we see that dissimilarity is generally higher when comparing two samples from the different locations than when comparing samples from the same location. We can use permutational analysis of variance (PERMANOVA) to test whether dissimilarity measured between different thallus locations is higher than dissimilarity measured between the same thallus locations. This is analagous to ordinary analysis of variance except that we must use permutations to determine whether the variance among groups is greater than the variance within groups because the distances between pairs of samples are not independent from one another. The `adonis()` function in the `vegan` package can perform this analysis. Just like `lm()` and `glm()`, the first argument is a formula relating the response variable to the predictor variable. In this case, the response variable is the object containing distances between samples calculated by `vegdist()`, and the predictor variable is the `Thallus_location` column from `tree1_data`. By default the function performs 999 permutations of the matrix to determine the probability of observing an F-ratio greater than or equal to the F-ratio from the observed data.

```{r}
# Conduct an analysis of variance on distance matrices
adonis(bc_dist ~ Thallus_location, data=tree1_data)
```

The model results show that the F-ratio is 2.73, which is greater than 1, indicating that dissimilarity between thallus locations is greater than dissimilarity within thallus locations. However, the permutation-based P-value is 0.1, which means that we would observe an F ratio this large (or larger) 10% of the time even if there were no difference in dissimilarity among thallus locations.


## Let's Review

**YOUR TURN**: For each function below, write a comment describing what it does

```{r}

lm()


glm()


rowSums()



predict()


vegdist()


```


