---
title: "Summarizing and Visualizing Data Part 1"
author: "BIO 46, Winter 2017, Stanford University"
date: "Week 3: January 24, 2017"
output: 
  html_document:
    highlight: haddock
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = '', eval=F)
```

## Lesson Goals

Today you will be learning how to manipulate data in R and do the following tasks:

1. Load a data table into R.
2. Subset and summarize data.
3. Combine two data tables.
4. Visualize data with boxplots.

Today we will work through the script named 'R-lesson2.r'. You should download this script from Canvas (Files/R-lessons/Code) and save it in your directory 'R-lessons/Code'. You can also download it from [GitHub](https://raw.githubusercontent.com/jescoyle/BIO46/master/R-lessons/Code/R-lesson2.r). As you work through this lesson, be sure to run each line of code in the script (which corresponds to the code in the lesson). If you can't tell how a whole line of code worked, try highlighting small pieces of the line of code and running them separately to see what they do. You only need to run each line of code once- any time that you assign an object a name, R will remember that object until you re-use that name for a different object. If you forget what objects you have created and named you can look in the Environment section of RStudio, which lists a description of each object you have created and assigned a name.

In the **YOUR TURN** sections you should type code that accomplishes the required task. When you complete the lesson, save your R script and upload it to the R Lesson 2 assignment on Canvas.

## Reading data tables into R

R has the capability to read data tables in many different text formats. The main type of data you will be usually be working with in R is a table where rows represent observations and columns describes attributes of those observations. R can read data tables stored as a text file with rows on separate lines and columns separated by a designated symbol, such as tabs, commas, or spaces. R cannot read Excel file formats directly. They must first be saved as a text file. Today we will be working with a text file where the columns are separated by commas. This type of file is called a *comma separated values* file, or *csv* file. 

You will be working with the file 'iButtons\_Fall2016.csv'. If you open this file in a text editor you can see how the data is stored, with commas separating each column. This data table contains temperature observations from iButtons deployed on 17 trees at Jasper Ridge. The iButtons recorded temperature in Celcius every 10 minutes from mid September to early October 2016. Each row in this file contains a temperature observation from an iButton deployed on a tree. Another file ('JRTrees\_Fall2016.csv') contains information about the trees where the iButtons where deployed.

Before beginning the lesson you should download each of these files from Canvas (Files/R-lessons/Data) and put them in the 'R-lessons/Data' directory on your compueter. 

There are two functions that allow you to read a csv file into R: `read.table()` is used for any text file. `read.csv()` is a short-cut function for csv files. Before you can use these functions you will first need to find the location where the data is stored, the *file path*. Use the File explorer (Windows) or Finder (Mac) to locate file path of the folder where you saved the data files you downloaded. You will use this information in the code below.

In the code below I first set an option that controls how text is interpreted when it is read into R. Then I define an object named `working_dir` which is a text string that tells R where to look for the data file. File paths in Windows start with a drive ('C:/') and for Macs they start with a '/'. In order to read the data on your computer you must change `working_dir` to the file path that is correct for your computer.

**YOUR TURN**: Change the working directory below so that you can read in the data tables from your 'R-lessons/Data' directory.

```{r read-in-data}
# Stop R from automatically converting text to categorical variables when reading in data
options(stringsAsFactors=F)

# Define the location of your working directory
working_dir = 'C:/Users/jrcoyle/Documents/Stanford/BIO46/GitHub/R-lessons/Data' # Windows

# Set the working directory for this session
setwd(working_dir)

# Read iButton data and save in a data frame names temp_data
# header = TRUE is used for data where the first row gives the column names
# sep =',' tells the function that columns are separated by commas
temp_data = read.table('iButtons_Fall2016.csv', header=TRUE, sep=',')

# Read tree data using the short-cut function, read.csv()
# By default this function assumes that column names are in the first row and that the columns are separated by commas
tree_data = read.csv('JRTrees_Fall2016.csv')
```


What do these data look like? In R studio you can use the Environment window to see that `tree_data` is a dataframe with 17 rows and 3 columns. There are also several functions that are useful for looking at the data in the console.

```{r}
# Look at the first few rows of tree_data
head(tree_data)

# Print out just the column names of tree_data
names(tree_data)

# Count the number of rows in tree_data
nrow(tree_data)

# Count the number of columns in tree_data
ncol(tree_data)

# Get the number of rows and columns together (e.g. the dimensions)
dim(tree_data)

```

**YOUR TURN**: Print out the column names of `temp_data` as well as the number of observations (each observation is stored in a row).

## Summarizing data fields

There are 17 different trees in tree_data. Let's check to see whether there are observations from each of these trees in the temperature data. The `table()` function counts the number of unique values present in a vector.

```{r}

# Display the TreeID column of temp_data
temp_data$TreeID

# Count the number of observations from each tree
table(temp_data$TreeID)

```

**Discuss with your partner**: Why does using `table()` on the TreeID column give the number of observations from each tree?

**YOUR TURN**: Make a vector named `species_count` that contains the number of trees measured for each tree species.

We can also summarize the temperature values.

```{r}
# What are the mean, minimum, and maximum temperature measured?
mean(temp_data$Temp)
min(temp_data$Temp)
max(temp_data$Temp)

# What is the range of temperature values?
range(temp_data$Temp)

```


## Arithmetic

When summarizing data we often want to make calculations. The most basic calculations that R can perform are simple arithmatic. Doing arithmatic is fairly straightforward-- we just write out the calculation that we want R to perform.


```{r arithmetic}
# Addition 
2 + 3 + 10

# Subtraction
6 - 2

# Division'
20 / 2

# Multiplication
10 * 3 * 2
10*3*2

# Exponents
3 ^ 2
3^2

# Order of operations
5 + 2 * 4^2
(5 + 2) * 4^2
5 + (2 * 4)^2

# Save the value of a calculation as an object named the_answer
the_answer = (12 / 2) * (10 - 3)

# Print the_answer
the_answer

```

Arithmatic can also be used on vectors or matrices. In this case, the calculation is performed element by element.

```{r vector-arithmetic}
# Select the first and last five temperature observations and save them as two vectors
first5 = temp_data[1:5, 'Temp']
last5_rows = (nrow(temp_data)-4):nrow(temp_data) # figure out which row numbers are the last 5 rows
last5 = temp_data[last5_rows, 'Temp']

# View them, then add them together
first5
last5
first5 + last5

# More arithmatic with vectors
last5 - 2
3 * last5^2

```

**YOUR TURN**: Add a column to `temp_data` called `temp_F` which is the Fahrenheit equivalent of the Celcius temperature displayed in `Temp`. Recall that you can use `temp_data$new_column_name = ` to add on a column called 'new\_column\_name' to a dataframe.

## Subsetting data

Suppose we only wanted to summarize temperature data from Tree 1. R uses logical operators to test whether certain conditions are true or false. These are: 


R Code   | Test
---------|---------------------------------
`A == B` | Is A equal to B?
`A < B`  | Is A less than B?
`A > B`  | Is A greater than B?
`A <= B` | Is A less than or equal to B?
`A >= B` | Is A greater than or equal to B?
`A != B` | Is A not equal to B?

So, in order to subset `temp_data` to only observations from Tree 1, first we need to test which rows correspond to Tree 1 and then use this vector of trues/falses to select the rows from Tree 1. You can select rows from a dataframe by using the square braces: `temp_data[get_these_rows, ]` or using the `subset()` function: `subset(temp_data, get_these_rows)`.

```{r}
# Find which rows of temp_data are from Tree 1
t1_rows = temp_data$TreeID == 'T1'
t1_rows

# How long is t1_rows?
length(t1_rows)

# How many rows are in temp_data?
nrow(temp_data)

# Is t1_rows the same length as temp_data?
length(t1_rows) == nrow(temp_data)

# Use the subset() function to select only rows from temp_data that correspond to when t1_rows is TRUE 
# e.g. the row is from tree 1
T1_data = subset(temp_data, t1_rows)

# Or, subset the data using square braces
T1_data = temp_data[t1_rows, ]

# Print out the dataframe to look at it
T1_data

```

Let's make a subset of the data when the iButtons recorded extreme temperatures. When you perform arithmetic with logicals (TRUE/FALSE), true counts as 1 and false counts as 0. This is can be useful for counting. 

```{r}
# Find observations when the recorded temperature was greater than 40 Celcius
high_temps = temp_data$Temp > 40

# How many temperature readings were greater than 40? (How many TRUES in high_temps?)
sum(high_temps)

# Which rows correspond to temperatures greater than 40? (Which elements in high_temps are TRUE?)
which(high_temps)

```

**YOUR TURN**:  

1. Make a dataframe called high\_temp\_data containing only observations where the temperature was greater than 40&deg;C.
2. Use the `table()` function to count the number of observations from each tree that had temperatures greater than 40&deg;C.

It isn't necessary to save a vector of logicals before subsetting. You can do it all in one step:

```{r}
# Find all observations where the temperature was less than 5 Celcius
# When using the subset function you can just use the name of the column
subset(temp_data, Temp < 5)

# If subseting using [], you must specify the actual column, not just its name:
temp_data[Temp < 5, ] # doesn't work because Temp doesn't exist as its own object
temp_data[temp_data$Temp < 5, ]

```

R allows you to combine logical statements using `&` (and) and `|` (or).

```{r}
# Find observations from Tree 3 where the temperature was greater than 35 Celcius
high_t3 = temp_data$TreeID == 'T3' & temp_data$Temp > 35
which(high_t3)

# Subset to these data using the logical vector high_t3
subset(temp_data, high_t3)

# Subset to these data without using a saved vector
subset(temp_data, TreeID == 'T3' & Temp > 35)

# Find observations where the temperature was greater than 40 or less than 5 Celcius
extreme_temps = temp_data$Temp > 40 | temp_data$Temp < 5


# Subset to these data (using the two different methods)
temp_data[extreme_temps, ] # method 1
temp_data[temp_data$Temp > 40 | temp_data$Temp < 5, ] # method 2


```

**YOUR TURN**:

1. How many temperature records were greater than 40&deg;C or less than 5&deg;C?
2. Make a dataframe called low\_t4 containing only observations from Tree 4 where the temperature was less than or equal to 10&deg;C.
3. How many temperature observations from Tree 4 were less than 5&deg;C?
4. Calculate the mean temperature from all data recorded at Tree 4. 

## Merging data tables

The tree data table shows that trees were located in two regions within Jasper Ridge, the north and the south. Suppose we wanted to know whether temperature differed between these regions. In order to use the data from `tree_data` to summarize data in `temp_data`, first we need to merge the two data sets by adding a column to `temp_data` containing the region where each temperature measurement was taken. The function to do this is `merge()`. We supply the two data frames as the first two agruemnts and then include a `by=` argument to tell the function which column is the same between the two dataframes.

```{r}
# Add the columns of tree_data onto temp_data based on TreeID
merged_data = merge(temp_data, tree_data, by='TreeID')

# Look at the first few rows of the new merged data
# Notice that all of the columns from tree_data have been added on to temp_data
head(merged_data)

# Calculate the mean temperature in the north and south regions
north_temps = subset(merged_data, Region == 'north')$Temp
mean(north_temps)

south_temps = subset(merged_data, Region == 'south')$Temp
mean(south_temps)
```

## Visualizing data with boxplots

Today we will visualize data in R using boxplots. Visualizing data is an important tool for evaluating whether data collected from different groups of observations differ from one another. First let's use a boxplot to look at temperature variation in the entire data set. In a boxplot, the middle line is the median of the data while the box extends to the 1st and 3rd quartiles (e.g., 50% of the observations are inside the box).

```{r}
# Make a boxplot of the temperature data from all trees
boxplot(temp_data$Temp)

```

We can make boxplots with data from each tree separately. To do this we will write a *formula* as the first argument. A formula tells R what to place on the y-axis and what to place on the x-axis. For example, to plot temperature (y-axis) as a function of tree identity (x-axis) the formula is: `Temp ~ TreeID`.

```{r}
# Make boxplots of temperature for each tree
boxplot(Temp ~ TreeID, data=temp_data)
```

Notice that we had to include the agrument `data=` to tell R where to look for the variable names `Temp` and `TreeID`. The plot needs axis labels. We can add these using the arguments `xlab` and `ylab`.

```{r}
# Make boxplots that include axis labels
boxplot(Temp ~ TreeID, data=temp_data, xlab='Tree', ylab=expression(Temperature~(degree*C)))
```

Finally, let's add a line for the median across the entire data set so that we can tell which trees have warmer and cooler temperatures. The function `abline()` tells R to add a line to an existing plot. The argument `h =` tells the function to add a horizontal line at the value following the `=`. 

```{r}
# Add red horizontal dashed line at the median of the entire data set
abline(h = median(temp_data$Temp), col='red', lty=2)
```

**YOUR TURN**: Make two boxplots to compare how temperature differs (1) between regions and (2) between tree species. Be sure to include axis labels on your plots. Hint: you may want to use the `merged_data` dataframe that we created earlier. **BONUS**: Use the `?` to figure out how to change the group labels below the boxes.


## Let's Review

**YOUR TURN**: For each of the following lines of code, write a comment above it describing (briefly) what it will do. Then, run the line of code to check your answer. Modify your comment if necessary.

```{r review}

jrtrees = read.table('JRTrees_Fall2016.csv', header=T, sep=',')


dim(jrtrees)


table(jrtrees$Region)


jrtrees[jrtrees$Region == 'north', ]


sum(jrtrees$Species == 'Quercus_agrifolia' & jrtrees$Region == 'north')

```