---
title: "Introduction to R"
author: "BIO 46, Winter 2017, Stanford University"
date: "Week 2: January 17, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = '')
knitr::opts_chunk$set(eval=F)
```

## Getting Started

R is a free software environment for data analysis and visualization. It has become a preferred tool among ecologists for analyzing research data. For this lesson you will need to download R (from: <http://r-project.og>) and install it. You should also download and install R Studio for desktop (from: <http://rstudio.com>).

The first step to success using R is to set up your files so that they are organized and easy to find. Before opening R, create a new directory ('folder') in the place on your computer where you store files for this class (maybe it's called 'BIO46'?). Name this new directory 'R-lessons'. Inside the 'R-lessons' directory create two more directories- one named 'Data', for data files, and one named 'Code', for files that will contain you analysis (these are called 'scripts'). We will use 'R_lessons' as your working directory in this and subsequent lessons.

Today we will work through the script named 'R-lesson1.r'. You should download this script from Canvas (Files/R-lessons/Code) and save it in your directory 'R-lessons/Code'. You can also download it from [GitHub](https://raw.githubusercontent.com/jescoyle/BIO46/master/R-lessons/Code/R-lesson1.r).

Now start R Studio and open 'R-lesson1.r'. In R Studio you can open scripts by clicking the folder icon or *File > Open File* and then navigating to the directory 'R-lessons/Code'. When the file opens, type in your name in the top line then click the save icon to save your work.

This file contains the code that you will see in this lesson document. There are two main parts- comments that give you information and code that actually does something:

```{r}
# Comments begin with a # and give you information about what the code does.
# For example, the next line is code that will print 'Hello' in the console.
print('Hello')
```

You will run (i.e. execute) each line of code in the interactive console. In R Studio, the console is the window in the lower left. It will have some text in it, and on the lowest line it should display:

`>`

If you see a 

`+`

then R is waiting for additional input. Make sure that `>` is displayed before you try to run each line of code. If your see `+` try pressing `Esc`. You should run each line of code in the script as you work through the lesson. You can run a line of code by placing your cursor on the line and either clicking the 'Run' button at the top right. You can run multiple lines by highlighting them and clicking 'Run'.

There will also be sections in the lesson where you will need to write your own code or add to an existing line. These will be marked by **YOUR TURN**. To complete the assignment you *must* do all of the **YOUR TURN** sections. You will save and submit your modified 'R-lesson1.r' script at the end of class. If you are stuck, ask your partner. If you are both stuck, ask for help- we are here to make sure everyone gets through the assignment. Occasionally there will be **BONUS** sections. These are not required, but there if you finish early and want a challenge. As you work through the document, be sure to click 'Save' periodically. R Studio does not have auto-save.

**YOUR TURN**: Modify the code above so that the console greets you by name.

## Objects

R works on things called 'objects'. An objects is anything saved in R that has a name and some information associated with that name. Today we will learn how to make and manipulate several kinds (i.e. *classes*) of objects: vectors, matrices, arrays, dataframes, and lists.

### Vectors

Vectors are a series of *elements* of a single *data type* (e.g. numbers or characters). The most basic way to create a vector is to use the concatenate function `c()`.

```{r vectors}
# This is a vector of numbers
c(1,2,3,4,5)

# This is a vector of characters
c('a','b','c','d','e')

# Vectors cannot contain elements with different types
c(1,2,3,'a','b')
```

**Discuss with your partner**: What happened in the third vector when you tried to combine three numbers and two letters?

Notice that each of these vectors printed to the console, but then where did they go? They no longer exists. If we want to *use* a vector then we have to give it a name so that it becomes an *object*. To retrieve the object, just type its name.

```{r vector-objects}
# Make a vector called fingers
fingers = c(1,2,3,4,5)

# Print the contents of lichens
fingers

# Make and view a vector called lichen
lichen = c('ascomycete','algae','yeast','bacteria')
lichen

```

**YOUR TURN**: Make a vector with the first 4 even numbers named 'evens'.

Here's a couple more ways to make vectors:

```{r making-vectors}
# Repeat something multiple times
rep('a', 5)
rep(evens, 2)
rep(evens, each=2)

# Make a sequence
1:10
20:15
seq(10)
seq(0, 10, by=2)
seq(0, 1, length.out=11)
seq(-10, 3, 0.5)

```

**YOUR TURN**: Make a vector named 'trials' with the integers from 1 to 50 repeated 3 times. E.g: 1,2,3,...,50,1,2,3,...,50,1,2,3,...50.

**Discuss with your partner**: How many elements should `trials` have in it?

### Matrices and Arrays

Matrices and arrays are just vectors with 2 (matrix) or more (array) dimensions. Like vectors, they must contain only one type of data. Here's how to make matrices and arrays:

```{r matrices-arrays}
# Make a 3 x 4 matrix filled of 0s
matrix(0, 3, 4)

# Make a 3 x 3 matrix with the integers 1 - 9
matrix(1:9, 3, 3)

# Make a 3-D array filled with cats
array('cat', dim = c(2,3,4))

# Make a matrix with more animals
animals = c('cat','parrot','fly','squid','seastar')
my_zoo = matrix(animals, 3, 5)

```

**Discuss with your partner:** 

* What happens when you change `dim = c(2,3,4)` to `dim = c(2,3,3)` when making that cat array?  
* What happens when you change it to `dim=c(2,3)`?  
* Predict which animal would be in the 4th row an 99th column of `matrix(animals, 5, 100)`.  

**YOUR TURN**: Make a matrix named trial_matrix with three columns, each of which contains the numbers 1 to 50 in increasing order.

### Lists and Dataframes

What if you want to combine elements that are different types of data? *Lists* are a series of elements that can have any data type. *Dataframes* are 2-dimensional tables where the elements within each column must have the same type of data, but different columns can hold different types of data.

Here are some examples:

```{r lists-dfs}

# Make some lists
list('My', 'dog', 'has', 10, 'fleas.')
list(c('My','dog','has'), 10, 'fleas.')
list(animals, my_zoo)
lichen_list = list('lichen', lichen)

# Make a dataframe
data.frame(1:4, lichen)
data.frame(part = 1:4, organism = lichen)

```

**Discuss with your partner**:

* Why are the two elements in lichen_list different?  
* What is the effect of including `part = ` before `1:4`?  

**YOUR TURN**:

* Add a vector containing the first four even numbers to the beginning of `lichen_list`.  
* Make a dataframe called 'hand' whose first column has the integers 1 to 5, second column contains finger names (starting with 'thumb'), and third column contains `animals`. Name the columns 'number','name','animal'.  

### Accessing elements in objects

Have you noticed the window in the upper right corner of R Studio? It should show the objects that you have currently defined in your environment. That is, it gives the names and contents of objects you have created. To show these objects in the console, type `ls()`.

Suppose you want to change the third object in `animals` to 'snail'. It would be very inconvenient to retype the whole vector. Instead R uses indices to access and change elements in objects.

```{r}
# Print the third element of animals
animals[3]

# Change the third element of animals to 'spider'
animals[3] = 'spider'
animals

```

To access and change elements in a list, use `[[ ]]`:

```{r}
# Print the first element in lichen_list
lichen_list[[1]]

# Change the first element in `lichen_list` to odd numbers
lichen_list[[1]] = seq(1, 7, by=2)

```

You can access multiple elements at once by using a vector of integers.

```{r}
# Get the first and last elements in lichen
lichen[c(1,4)]

# Display the elements of lichen in reverse the order
lichen[4:1]

```

**YOUR TURN**: Create a vector named 'little_zoo' containing only parrots, squids, seastars and algae without typing the names of any of these animals.

You can access elements in objects with 2 or more dimensions by separating indices with commas.

```{r}
# Which animal is in the 2nd row and 4th column of my_zoo?
my_zoo[2,4]

# Which animals are in the 3rd row of my_zoo?
my_zoo[3,]

# Which animals are in the 4th and 5th columns of my_zoo?
my_zoo[,c(4,5)]

```

**Discuss with your partner**:

* What happens when you type `my_zoo[3]`? What about `my_zoo[5]`?  
* What happens when you type `lichen[2,3]`? Why? How would you change this to access the 2nd and 3rd elements of `lichen`?  

**BONUS**: Change the last two odd numbers in `lichen_list` back to even numbers.

We can also access elements in objects using their names. For example, when you created the dataframe `hand` you gave each of its columns names. Other objects, like vectors, lists and matrices can have names for their elements, rows and columns.

```{r element-names}
# Name the rows and columns of a matrix
rownames(my_zoo) = c('A','B','C')
colnames(my_zoo) = c('blue','pink','red','yellow','green')
my_zoo

# Get the animals in the blue column in rows A and B
my_zoo[c('A','B'), 'blue']

# Create a data frame
my_data = data.frame(1:20, lichen, rep(1:5, each=4))

# Change the column names
names(my_data) = c('trial', 'organism', 'treatment')
names(my_data)
colnames(my_data)
my_data

# Name the elements in a vector
names(animals) = c('Felix', 'Polly', 'Charlotte', 'Kraken', 'Patrick')

# What type of animal is Patrick?
animals['Patrick']

# Name the elements in a list
names(lichen_list) = c('numbers','organisms')
lichen_list

# Access the elements of a list by name
lichen_list[['organisms']]
lichen_list[['organisms']][2]

```

**YOUR TURN**: Extract the organisms associated with odd trial numbers in `my_data`.

You can also extract and modify columns in a dataframe using a `$`: `dataframe_name$column_name`. This makes it easy to add new columns to an existing dataframes.

```{r dataframes}
# Display the treatment column of my_data
my_data$treatment

# Add a new column to my_data called temp
my_data$temp_C = rep(c(5,10,15,20,25), each=4)
my_data

```


## Functions

We will use functions to create and modify objects in order to perform data analyses. Each function has a name and a set of arguments that tell it how to behave:

`function_name(argument1, argument2, argument3)`

You have already used several functions in this lesson, for example: `seq()`, `rep()`, `c()`. To see what a function does, just type the function name:

```{r}
# Display a function
seq
```

Well, that wasn't very informative. To learn how to use how to use a function you can type `?function_name`. In R Studio this will display a help file in the lower right window.

```{r}
# Display the help file on seq()
?seq

```

The *Arguments* section of the help file shows the names of the arguments that you can use. For example in the help file for `seq()` the arguments are `from`, `to`, `by`, `length.out`, `along.with`. This section also shows the values that each argument will take if you do not give it a value. The *Usage* sections shows the order of these arguments. If you use a function with its arguments in order, you do not need to give the actual names. However, if you *do not* put the arguments in the correct order, you *must* use the names.

```{r arguments}
# Make a sequence of numbers from 10 to 20 counting by 5
seq(10, 20, 5)

# Make a sequence of 6 numbers spread evenly between 10 and 20 
seq(10, 20, length.out=6)

```

**YOUR TURN**: Display the help file on `matrix()`. Make a matrix named row_matrix that has 4 rows, each with the numbers 1 to 5, in order.

### Arithmetic 

We can also use functions to do arithmetic on numeric objects. These functions are *operators*, they don't use parentheses with arguments inside. Instead we just write out the calculation that we want R to perform.

```{r arithmetic}
# Add two numbers
2+3

# Add the first two elements of fingers
fingers[1] + fingers[2]

# Subtraction
monkey = 6 - 2
monkey

# Division'
monkey / 2

# Multiplication
monkey*3

# Exponents
3^2

# Order of operations
5+2*4^2
(5+2)*4^2
5+(2*4)^2

```

These operations can also be used on vectors or matrices. In this case, the calculation is performed element by element.

```{r vector-arithmetic}
# Create two vectors
troupe = rep(monkey, 7)
troupe
bananas = seq(-3,3)
bananas

# Arithmetic with vectors and scalars
bananas + 3
bananas * 2
bananas^2

# Vector arithmetic
troupe + bananas
troupe - bananas
troupe * bananas
troupe / bananas
troupe^bananas

```

***YOUR TURN**: Add a column to `my_data` called `temp_F` which is the Fahrenheit equivalent of the Celcius temperature displayed in `temp_C`.

## Let's Review

**YOUR TURN**: For each of the following lines of code, write a comment above it describing (briefly) what it will do. Then, run the line of code to check your answer. Modify your comment if necessary.

```{r review}

fox = 2:8


c(fox, 2)


rep(fox, 3)


seq(0, 10, 2)


matrix(1:9, nrow=3)


array(fox, dim=c(2,2,4))


list('myfox', fox)


fox_data = data.frame(fox, rep('hare', 7))


names(fox_data)[2] = 'prey'


fox_data$kits = fox*2

```


