---
title: "Summarizing and Visualizing Data Part 2"
author: "BIO 46, Winter 2017, Stanford University"
date: "Week 4: January 31, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = '', eval=F)
```

## Lesson Goals

Out goal today is to use the temperature data recorded by iButtons on trees at Jasper Ridge in the fall of 2016 to determine how temperature varies daily and across different locations. On Thursday we will be starting a 3-week growth chamber experiment and will use the information from today's lesson to decide how to program the temperature in the growth chamber.

We will be working with the same two data files from the last lesson: 'iButtons\_Fall2016.csv' and 'JRTrees\_Fall2016.csv'. Be sure that these files are downloaded and saved in a folder on your computer. Today we will work through the script named 'R-lesson3.r'. You should download this script from Canvas (Files/R-lessons/Code) and save it in your directory 'R-lessons/Code'. You can also download it from [GitHub](https://raw.githubusercontent.com/jescoyle/BIO46/master/R-lessons/Code/R-lesson3.r).

While working through this script you will learn how to:

1. Make a bivariate plot (an x-y plot).
2. Convert text to dates and times.
3. Summarize data using functions in the dplyr package.

As you work through this lesson, be sure to run each line of code in the script (which corresponds to the code in the lesson). Remember- lines that begin with `#` are comments that tell you what the next line of code is supposed to do.
If you can't tell how a whole line of code worked, try highlighting small pieces of the line of code and running them separately to see what they do. You only need to run each line of code once- any time that you assign an object a name, R will remember that object until you re-use that name for a different object. If you forget what objects you have created and named you can look in the Environment section of RStudio, which lists a description of each object you have created and assigned a name.

In the **YOUR TURN** sections you should type code that accomplishes the required task. You do not need to change any code that is not in the **YOUR TURN** sections. If you get stuck on a task, move on and come back to it later. The **YOUR TURN** tasks do not depend on one another. When you complete the lesson, save your R script and upload it to the R Lesson 3 assignment on Canvas. 

### Task 1: Plot variation in temperature at one tree during Fall 2016

First we need to read the data into a dataframe table that we can work with in R.

```{r read-in-data}
# Stop R from automatically converting text to categorical variables when reading in data
options(stringsAsFactors=F)

# Define the location of your working directory
working_dir = 'C:/Users/jrcoyle/Documents/Stanford/BIO46/GitHub/R-lessons/Data' # Windows

# Set the working directory for this session
setwd(working_dir)

# Read iButton data and save in a data frame names temp_data
# header = TRUE is used for data where the first row gives the column names
# sep =',' tells the function that columns are separated by commas
temp_data = read.table('iButtons_Fall2016.csv', header=TRUE, sep=',')

# Read tree data using the short-cut function, read.csv()
# By default this function assumes that column names are in the first row and that the columns are separated by commas
tree_data = read.csv('JRTrees_Fall2016.csv')
```

**YOUR TURN**: Change the working directory in the code above so that the code will find the files you saved on your computer.

Next, let's chose a Coast live oak (*Quercus agrifolia*) from the southern part of Jasper Ridge (which is where the trees we sample for class are located) and subset the temperature data to only contain data recorded at that tree.

```{r subset-to-tree}
# Look at the contents of the tree_data dataframe
tree_data

# Let's choose Tree 13 and subset the temperature data to just this tree
focal_tree = 'T13'
focal_tree_data = subset(temp_data, TreeID==focal_tree)

# Look at the first five rows of focal_tree_data
head(focal_tree_data)

# How many temperature observations are there?
nrow(focal_tree_data)
```

We want to display a time series of the temperature values recorded at Tree 13. The column of `focal_tree_data` that contains the information about when a temperature was recorded is called `Date_time`. However, at the moment R doesn't know that this column contains dates, it only sees the data as text (aka 'character'):

```{r}
# Display the type of data stored in the Date_time column of focal_tree_data
class(focal_tree_data$Date_time)
```

We need to convert this column to a format that R understands as dates. The function to do this is `strptime()`. The first argument we give this function is the vector of text-based dates. The second argument is a text string the tells the function the format that the dates are in. Don't worrry about the details here. You will not need to modify this code.

```{r strptime}
# Make a vector of time points that corresponds to Date_time and is in a date format that R can understand
time_points = strptime(focal_tree_data$Date_time, format='%m/%d/%y %I:%M:%S %p %z')

# View these new times
time_points

# Display the type of data stored in the time_points vector - it is not 'character'
class(time_points)
```

Now that R is interpreting the `time_points` vector as a date, we can plot temperature versus time to see the time-series of temperature values at Tree 13. The function to make a bivariate plot (x-y plot) is called `plot()`. The most basic way to use this function is to put the x values in the first argument and the y values in the second argument: `plot(xvalues, yvalues)`. To make our plot the x values are in the `time_points` vector and the y values are in the 'Temp' column of `focal_tree_data`.


```{r plot_time_series}
# Plot temperature versus time using method 1:
plot(time_points, focal_tree_data$Temp)
```

There are many ways we can improve this plot by adding arguments. First, notice that each observation was plotted as a point. Since this is a time-series we probably want to connect the points in a single line. The argument to do this is `type='l'`, where the 'l' tells the function to make a line. Second, we should re-label the axes to make them more interpretable. The arguments to do this are `xlab='x-axis name'` and `ylab='y-axis name'`. Finally, it is good practice to have the y-axis number be oriented vertically. This is done usin the argument `las=1`.

```{r better_time_series}
# Plot temperature vs time as a line and with axis labels
plot(time_points, focal_tree_data$Temp, type='l', xlab='Date', ylab='Temperature (C)', las=1)
```

Can you see the daily temperature cycles? Notice how the amplitude of the cycles changes on September 20th. This was the date when the iButton was placed outdoors on Tree 13. In order to analyze the data we should remove all observations before this date.

```{r}
# Define the cut-off date to remove data
cutoff_date = strptime('2016-09-20 12:00:00 -0700', format='%Y-%m-%d %H:%M:%S %z')

# Which time_points are after the cutoff date?
time_points > cutoff_date

# Subset focal_tree_data to only observations after the cutoff date
analysis_data = subset(focal_tree_data, time_points > cutoff_date)

# How many observations remain for analysis?
nrow(analysis_data)
```

**YOUR TURN**: Calculate the number of temperature observations that were dropped from the analysis because they occured prior to the cutoff date.

### Task 2: Summarize the temperature data at one tree

Now that we have a dataframe containing only temperature observations from a single tree we can use R to summarize the temperature enviornment. First, let's re-plot the temperature time-series and add horizontal lines at the minimum and maximum temperatures recorded.

```{r}
# Make a new vector of time_points that corresponds to the times in analysis_data
analysis_times = strptime(analysis_data$Date_time, format='%m/%d/%y %I:%M:%S %p %z')


# Plot temperature vs time as a line and with axis labels
plot(analysis_times, analysis_data$Temp, type='l', xlab='Date', ylab='Temperature (C)', las=1)

# Add a red horizontal line at the maximum temperature measured
abline(h = max(analysis_data$Temp), col='red')

# Add a blue horizontal line at the minimum temperature measured
abline(h = min(analysis_data$Temp), col='blue')
```

**YOUR TURN**: Add a purple horizontal line to the plot at the mean temperature value. Hint: the function for calculating the mean of a vector is `mean()`.

### Task 3: Summarize daily temperature data at one tree

Suppose we want to know not just the overall minimum and maximum temperatures, but the daily high and low temperatures. In order to do this calculation we need to assign each temperature observation to a date and then summarize temperature measurements by date.

```{r}
# Make a vector that gives the date of each temperature measurement without the time.
analysis_dates = as.Date(analysis_times)

# Plot the temperature measurements by date
plot(analysis_dates, analysis_data$Temp, xlab='Date', ylab='Temperature (C)', las=1)
```

We will use functions from the dplyr package to summarize temperatures by date. This package is not automatically included when you download R, so we will need to install the package and then load it to use the functions in it.

To install a package in R Studio you can click the 'Packages' tab in the upper right window, then click 'Install' and type 'dplyr' into the line marked 'Packages'. When you click install watch the console. Notice that `install.packages('dplyr')` appears. You could have also just typed this directly into the console to download and install the package.

You only need to install a package once, but in order to use the functions in a package you need to load the functions each time you start R.

```{r}
# Load the package dplyr
library(dplyr)
```

The functions in the dplyr can do a lot and we will learn just a little today. If you want to learn more there is a great introduction [here](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html). 

The `summarise()` functions allows us to make several calculations on a dataframe at once and store the results as a new vector. We can name these calculations to make the results easier to interpret. For example, in the code below I calculate the number of rows in `analysis_data` and name this value `Num_obs`. Then I calculate the mean of the `Temp` column in `analysis_data` and name the result `avgT`.


```{r}
# Count the number of observations in analysis_data
summarise(analysis_data, Num_obs=n())

# Calculate the average temperature 
summarise(analysis_data, avgT = mean(Temp))

# Do both at once
summarise(analysis_data, Num_obs = n(), avgT = mean(Temp))
```

In the example above, the `summarise()` function is working on all of the data at once. In order to summarize temperatures on each date we first need to create a grouped table using the `group_by()` function. Then, when we use `summarise()` it will automatically give results for each date.

```{r}
# Add the dates as a column in the analysis_data dataframe
analysis_data$Date = analysis_dates

# Look at the first few rows to see that it was added
head(analysis_data)

# Create a table grouping temperature observations by date
grouped_data = group_by(analysis_data, Date)

# Summarize temperature by date and save as a new dataframe called daily_temp_summary
daily_temp_summary = summarise(grouped_data, Num_obs=n(), avgT=mean(Temp), maxT=max(Temp))

# View the summary data
daily_temp_summary

```

**YOUR TURN**: Modify the code that creates the `daily_temp_summary` dataframe so that it includes a column called 'minT' containing the minimum temperature measured each day.

We can calculate average daily minimum and maximum temperatures using the `daily_temp_summary` dataframe we created above.

```{r}
# Calculate the mean daily high temperature at this tree
mean(daily_temp_summary$maxT)

```

**YOUR TURN**: Calculate the mean daily low temperature at this tree

### Task 4: Summarize daily temperature measurements at multiple trees

If you look at the `tree_data` dataframe you will see that there are seventeen trees. We want to know what the average daily high and low temperatures are at each of these trees, not just Tree 13. We could look at each tree in turn by going back up to the top of this R script and changing `focal_tree` to `'T14'`, `'T15'`, etc. and then re-running each line of code. 

**YOUR TURN**: Change the `focal_tree` at the beginning of this R script so that subsequent code will analyze data from Tree 17. What was the mean daily maximum temperature at Tree 17? (Write your answer, don't copy down all of the code again.)

Clearly, it would be tedious to have to change the `focal_tree` and re-run the script 5 times. Instead, let's group the temperature data by tree AND day, then summarize it. 


```{r}
# Make a vector of time points that corresponds to Date_time and is in a date format that R can understand
time_points = strptime(temp_data$Date_time, format='%m/%d/%y %I:%M:%S %p %z')

# Subset temp_data to only observations occuring after the cutoff date
analysis_data = subset(temp_data, time_points > cutoff_date)

# Make a new vector of time_points that corresponds to the times in analysis_data
analysis_times = strptime(analysis_data$Date_time, format='%m/%d/%y %I:%M:%S %p %z')

# Extract the dates from analysis_times and add them as a new column in analysis_data
analysis_data$Date = as.Date(analysis_times)

# Group analysis_data by TreeID and by Date
grouped_data = group_by(analysis_data, TreeID, Date)

# Calculate the mean and max daily temperatures at each tree
temp_summary = summarise(grouped_data, avgT=mean(Temp), maxT=max(Temp))

# Display temp_summary
temp_summary

# Convert temp_summary to a regular dataframe so that all rows display
temp_summary_df = as.data.frame(temp_summary)
temp_summary_df

# View summary data from one tree
subset(temp_summary_df, TreeID=='T14')

# Calculate the average daily high temperature across all trees
mean(temp_summary_df$maxT)

```
**YOUR TURN**: Calculate the average daily low temperature across all trees.


## Let's Review

**YOUR TURN**: For each line of code below, write a comment above it describing what it does. You can assume that `temp_summary_df` is the same dataframe we just created in the code above, containing average and maximum daily temperatures for each tree on each day.

```{r}

plot(temp_summary_df$Date, temp_summary_df$maxT)


plot(temp_summary_df$avgT, temp_summary_df$maxT, xlab='Mean temp.', ylab='Max. temp.')


abline(h=mean(temp_summary_df$maxT))


subset(temp_summary_df, maxT > 40)


tree_groups = group_by(temp_summary_df, TreeID)


summarise(tree_groups, mean_high=mean(maxT))
```
