---
title: "Extra Code"
author: "BIO 46, Winter 2017, Stanford University"
date: "January 27, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




## Merging data tables

The tree data table shows that trees were located in two regions within Jasper Ridge, the north and the south. Suppose we wanted to know whether temperature differed between these regions. In order to use the data from `tree_data` to summarize data in `temp_data`, first we need to merge the two data sets by adding a column to `temp_data` containing the region where each temperature measurement was taken. The function to do this is `merge()`. We supply the two data frames as the first two agruemnts and then include a `by=` argument to tell the function which column is the same between the two dataframes.

```{r}
# Add the columns of tree_data onto temp_data based on TreeID
merged_data = merge(temp_data, tree_data, by='TreeID')

# Look at the first few rows of the new merged data
# Notice that all of the columns from tree_data have been added on to temp_data
head(merged_data)

# Summarize temperature by region
temp_byReg = group_by(merged_data, Region)
reg_summary = summarise(temp_byReg, count = n(), avgT = mean(Temp), highT = max(Temp), lowT = min(Temp))
reg_summary

```

**Discuss with your partner**: Is one region cooler than the other? Why or why not?

**YOUR TURN**: Summarize temperature differences between the two tree species. Name this summary table `species_summary`.

