---
title: "Introduction to R"
author: "Jes Coyle"
date: "October 28, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting Started

R is a free software environment for data analysis and visualization. It has become a preferred tool among ecologists for analyzing research data. For this lesson you will need to download R (from: <http://r-project.og>) and install it. You should also download and install R Studio for desktop (from: <http://rstudio.com>).

The first step to success using R is to set up your files so that they are organized and easy to find. Before opening R, create a new directory ('folder') in the place on your computer where you store files for this class (maybe its called 'BIO46'?). Name this new directory 'R_lessons'. Inside the 'R_lessons' directory create two more directories- one named 'Data', for data files, and one named 'Code', for files that will contain you analysis (these are called 'scripts'). We will use 'R_lessons' as your working directory in this and subsequent lessons.

Today we will work through the script named 'R-lesson1.r'. You should download this script from Axess or obtain it from <http://github.com/jescoyle/BIO46/R_lessons/Code/R-lesson1.r>

Now open R Studio.





```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
